WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-04-15 ms.tex
Use acronyms in environment
Add stuff from thesis

\documentclass[amsmath,amssymb,aps,prx,reprint,nopreprintnumbers,nofootinbib]{revtex4-1}
\usepackage{multirow}
\usepackage{bigdelim}
\usepackage{multirow}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{bib/aas_macros}
\usepackage[colorlinks]{hyperref}
\usepackage{protosem}
\input{ligo-acronyms/acronyms}

@@

\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\std}{std}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}

@@

@@

\begin{abstract}
Within the next few years, Advanced LIGO and Virgo should detect gravitational waves from binary neutron star and neutron star\nobreakdashes--black hole mergers. These sources are also predicted to power a broad array of electromagnetic transients. Because the X\nobreakdashes-ray and optical signatures can be faint and fade rapidly, observing them hinges on rapidly inferring the sky location from the gravitational wave observations. MCMC methods for gravitational\nobreakdashes-wave parameter estimation can take hours or more. We introduce BAYESTAR, a rapid, Bayesian, non-MCMC sky localization algorithm that takes just seconds to produce probability sky maps that are comparable in accuracy to the full analysis. Prompt localizations from BAYESTAR will make it possible to search electromagnetic counterparts of compact binary mergers.
\end{abstract}

@@

Ground-based \ac{GW} interferometers are entering the advanced detector era, positioning themselves to make profound discoveries about the universe. The \ac{aLIGO} and Vigo detectors will be part of a worldwide network that also includes KAGRA and hopefully LIGO-India. Coalescing neutron star binaries are among the most likely sources, with 0.4\nobreakdashes--400 detections per year~\cite{rates}. In addition to being an efficient source of \acp{GW}, the \ac{NS} may be tidally shredded before merging, providing fuel for an electromagnetic counterpart. The short-lived accretion flow may power a collimated relativistic jet, resulting in a short \ac{GRB} and an X-ray afterglow if it is aligned with the line of sight. An optical afterglow may follow minutes to hours later as the jet plows into the interstellar medium. As surrounding neutron-rich ejecta decay radioactively, an optical `kilonova' may be visible after $\sim$1 day \cite{metzger:2010}. Finally, as the ejecta plow into the interstellar medium, a faint radio afterglow could occur on a timescale of weeks to years \cite{Nakar:2011cw}.

There is therefore a strong case for performing electromagnetic followups of these sources of \acp{GW}. Indeed, the final science run of the intitial LIGO and Virgo instruments saw the first joint search for \acp{GW} from compact binaries and their electromagnetic counterparts \cite{abadie2012first}.  As we prepare for the next generation of these searches there is a need for determining the location of the source as precisely as possible in as little time as possible.

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques in several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter space we need to sample. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \ac{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}%
. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

\section{Theory of \ac{GW} detection and parameter estimation}

In the frequency domain, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}\label{eq:signal-model}
    y_i (\omega) = x_i (\omega; \bm\theta) + n_i (\omega),
\end{equation}
%
where $x_i (\omega; \bm\theta)$ is the \ac{GW} signal given a parameter vector $\bm\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[\left|n_i(\omega)\right|^2\right] + E\left[\left|n_i(-\omega)\right|^2\right] = 2 E\left[\left|n_i(\omega)\right|^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf y (\omega) \equiv \{y_i (\omega)\}_i$.

The likelihood, or the probability of the observation $\mathbf y$ conditioned on the value of $\bm\theta$, is a product of Gaussian distributions:
%
\begin{multline}\label{eq:gaussian-likelihood}
    \mathcal{L}(\mathbf y; \bm\theta) = \prod_i p(y_i | \bm\theta)
        \\ \propto \exp \left[
        - \frac{1}{2} \sum_i \int_0^\infty \frac{\left|y_i (\omega)
            - x_i(\omega; \bm\theta) \right|^2}{S_i(\omega)} \, d\omega
    \right].
\end{multline}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \bm\theta = \left[
    \begin{array}{cl@{\quad}p{4cm}lp{3.75cm}}
        \alpha & \rdelim]{11}{0mm} & right ascension & \rdelim\}{7}{1mm} & \multirow{7}{2cm}{extrinsic parameters, $\bm\theta_\mathrm{ex}$} \\
        \delta && declination & \\
        r && distance & \\
        t_\oplus && arrival time at geocenter & \\
        \iota && inclination angle & \\
        \psi && polarization angle & \\
        \phi_c && coalescence phase & \\
        \cline{1-1}\cline{3-3}
        m_1 && first component's mass & \rdelim\}{4}{1mm} & \multirow{4}{2cm}{intrinsic parameters, $\bm\theta_\mathrm{in}$.}\\
        m_2 && second component's mass & \\
        \mathbf S_1 && first component's spin & \\
        \mathbf S_2 && second component's spin & \\
    \end{array}\right.
\end{equation}
%
This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it does play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\cite{PhysRevD.87.043004}, \ac{BNS} systems formed by binary stellar evolution should circularize due to tidal interaction~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$.

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h = h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. If $h_0$ and $h_{\pi/2}$ are Fourier transforms of real functions, then $h_0(\omega) = h_0^*(-\omega)$ and $h_{\pi/2}(\omega) = h_{\pi/2}^*(-\omega)$, and we can write (assuming an arbitrary phase convention)
%
\begin{equation}
    h_{\pi/2}(\omega) = h_0(\omega) \cdot
    \begin{cases}
        -i & \text{if } \omega \geq 0 \\
        i & \text{if } \omega < 0
    \end{cases}.
\end{equation}
%
Then, we can write the signal model in a way that isolates all dependence on the extrinsic parameters $\bm\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\bm\theta_\mathrm{in}$ into the basis waveform, by taking the Fourier transform of Equation~(2.8) of \cite{PhysRevD.83.084002}:
%
\begin{multline}\label{eq:full-signal-model}
    x_i(\omega; \bm\theta) = e^{-i \omega t_\oplus}
    \frac{r_{1,i}}{r}
    e^{2 i \phi_c} \\
    \left[
    \frac{1}{2} \left(1 + \cos^2 \iota\right) \Re \left\{\zeta\right\} - i
    \left(\cos\iota\right) \Im \left\{\zeta\right\}
    \right]
    h(\omega; \bm\theta_\mathrm{in})
\end{multline}
%
for $\omega \geq 0$, where
%
\begin{equation}
    \zeta = e^{-2 i \psi} \left(
    F_{+,i}(\alpha, \delta, t_\oplus) +
    i F_{\times,i}(\alpha, \delta, t_\oplus)
    \right).
\end{equation}
%
The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, defined such that $0 \leq {F_{+,i}}^2 + {F_{\times,i}} \leq 1$. They depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event, and are derived in~\cite{PhysRevD.63.042003}. The quantity $r_{1,i}$ is a fiducial distance at which detector $i$ would register SNR=1 for an optimally oriented binary (face\nobreakdashes-on, and in a direction perpendicular to the interferometer's arms),
%
\marginpar{I seem to have dropped a factor of 2 here...}%
%
\begin{equation}\label{eq:horizon}
r_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,d\omega.
\end{equation}

More succinctly, we can write the signal received by detector $i$ in terms of observable extrinsic parameters $\bm\theta_i = (\rho_i, \gamma_i, \tau_i)$, the amplitude $\rho_i$, phase $\gamma_i$, and time delay $\tau_i$ on arrival at detector $i$,
%
\begin{equation}\label{eq:signal-model}
    x_i (\omega; \bm\theta_i, \bm\theta_\mathrm{in}) = x_i (\omega; \rho_i, \gamma_i, \tau_i, \bm\theta_\mathrm{in}) = \frac{\rho_i}{\sigma_i} e^{i (\gamma_i - \omega \tau_i)} h(\omega; \bm\theta_\mathrm{in}).
\end{equation}

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from the likelihood in Equation~(\ref{eq:gaussian-likelihood}) and the signal model in Equation~(\ref{eq:signal-model}). Concretely, this results in a bank of matched filters, or the cross-correlation between a template waveform and the incoming data stream,
%
\begin{equation}
z_i(\tau_i;\bm\theta_\mathrm{in}) = \frac{1}{\sigma_i (\bm\theta_\mathrm{in})} \int_0^\infty \frac{h^*(\omega; \bm\theta_\mathrm{in}) y_i(\omega) e^{i \omega \tau_i}}{S_i(\omega)} \,d\omega.
\end{equation}
%
The \ac{ML} point estimates of the signal parameters, $\mathrm{MLE}(\mathbf{y}) = \{\{ \hat{\bm\theta}_i \}_i, \hat{\bm\theta}_\mathrm{in}\} = \{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\bm\theta}_\mathrm{in}\}$, are given by
%
\begin{eqnarray}
    \hat{\bm\theta}_\mathrm{in}, \{\hat\tau_i\}_i
        &=& \argmax_{\bm\theta_\mathrm{in}, \{\hat\tau_i\}_i}
        \sum_i \left| z_i\left(\tau_i;
        \bm\theta_\mathrm{in}\right) \right|^2, \\
    \hat\rho_i &=& \left| z_i\left(\hat\tau_i;
        \hat{\bm\theta}_\mathrm{in}\right) \right|, \\
    \hat\gamma_i &=& \arg z_i\left(\hat\tau_i;
        \hat{\bm\theta}_\mathrm{in}\right).
\end{eqnarray}
%
A detection candidate consists of $\{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\bm\theta}_\mathrm{in}\}$. There are various ways to characterize the significance of a detection candidate, but in Gaussian noise the combined SNR $\hat\rho_\mathrm{c}$ suffices,
%
\marginpar{What is the least that I can get away with saying about false alarm rate estimation and likelihood ratios?}%
%
\begin{equation}
    \hat\rho_\mathrm{c} = \sqrt{\sum_i {\hat\rho_i}^2}.
\end{equation}

\subsection{Bayesian probability and parameter estimation}

In the Bayesian framework, the parameters are inferred from the data by forming the posterior distribution, $p(\bm\theta|\mathbf y)$, which describes the probability of the parameters given the observations. Bayes' rule relates the likelihood $p(\mathbf y|\bm\theta)$ to the posterior $p(\bm\theta|\mathbf y)$,
%
\begin{equation}\label{bayes}
p(\bm\theta|\mathbf y) = \frac{p(\mathbf y|\bm\theta) p(\bm\theta)}{p(\mathbf y)},
\end{equation}
%
introducing the prior distribution $p(\bm\theta)$ which encapsulates previous information about the paramters (for example, arising from earlier observations or from known physical bounds on the parameters), and the evidence $p(\mathbf y)$ which can be thought of as a normalization factor or as describing the parsimoniousness of the model.

The choice of prior is open to one's astrophysical preconceptions, but during \ac{S6} when \ac{LIGO}'s Bayesian \ac{CBC} parameter estimation pipelines were pioneered, the prior was taken to be isotropic in sky location and binary orientation, and uniform in volume, arrival time, and the component masses~\cite{2013arXiv1304.1775T}.

>>>>

\documentclass[amsmath,amssymb,aps,prx,reprint,nopreprintnumbers,nofootinbib]{revtex4-1}
\usepackage[nolist]{acronym}
\usepackage{multirow}
\usepackage{bigdelim}
\usepackage{multirow}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{bib/aas_macros}
\usepackage[colorlinks]{hyperref}
\usepackage{protosem}

@@

\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\std}{std}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}

@@

\newcommand\transpose{\ensuremath{^{^\mathsf{T}}}}

@@

\begin{abstract}
Within the next few years, Advanced LIGO and Virgo should detect \acp{GW} from binary neutron star and neutron star\nobreakdashes--black hole mergers. These sources are also predicted to power a broad array of electromagnetic transients. Because the X\nobreakdashes-ray and optical signatures can be faint and fade rapidly, observing them hinges on rapidly inferring the sky location from the gravitational wave observations. MCMC methods for gravitational\nobreakdashes-wave parameter estimation can take hours or more. We introduce BAYESTAR, a rapid, Bayesian, non-MCMC sky localization algorithm that takes just seconds to produce probability sky maps that are comparable in accuracy to the full analysis. Prompt localizations from BAYESTAR will make it possible to search electromagnetic counterparts of compact binary mergers.
\end{abstract}

@@

The \acf{LIGO} and Virgo will soon begin taking data \cite{LIGOObservingScenarios} in their `Advanced' configurations, ultimately attaining an order of magnitude more sensitivity than in their initial configurations through 2010. They will form the first parts of a sensitive global \ac{GW} detector network, to be later augmented by the Japanese KAGRA facility \cite{LCGT,KAGRAInterferometerDesign} and \ac{LIGO}\nobreakdashes--India \cite{LIGOIndia}.

The most readily detectable sources of \acp{GW} include binary neutron star mergers, with 0.4\nobreakdashes--400 events per year within the reach of Advanced LIGO at final design sensitivity~\cite{rates}. These binary systems are not only efficient sources of \acp{GW}, but also potential sources of detectable \ac{EM} transients from the aftermath of the tidal disruption of the \acp{NS}. \citet{MostPromisingEMCounterpart} argue that the most promising \ac{EM} counterparts are the hypothesized optical/infrared ``kilonovae'' powered by the radioactivate decay of $r$\nobreakdashes-process elements synthesized within the neutron\nobreakdashes-rich ejecta. These are expected to be faint, red, and peak rapidly, reaching only $R \sim 24$~mag within a week, though rising several magnitudes brighter in the infrared \citep{KilonovaHighOpacities}.

Several mechanisms could make the kilonovae brighter, bluer, and hence more readily detectable \citep{KilonovaPrecursor,KilonovaRedOrBlue}, but peak even earlier, within hours. If, as is widely believed \citep{1986ApJ...308L..43P,1989Natur.340..126E,1992ApJ...395L..83N,2011ApJ...732L...6R}, \ac{BNS} mergers are indeed progenitors of short \acp{GRB}, then a small (due to jet collimation) fraction of Advanced \ac{LIGO} events could also be accompanied by a bright optical afterglow, but this signature, likewise, would peak within hours or earlier.

Adding to the challenge of detecting a faint, short-lived optical transient, \ac{GW} detector networks have inherently coarse position reconstruction accurate to tens to hundreds of deg$^2$ \citep{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}. Wide\nobreakdashes-field optical transient surveys such as the \acf{ZTF}, BlackGEM, or the \acf{LSST}, operated in ``target of opportunity'' mode, may be able to tile these large areas rapidly enough to find the one needle in the haystack that is connected to the \ac{GW} event, but clearly time will be of the utmost.

The final science run of the initial LIGO and Virgo instruments saw the first joint search for \ac{GW} and electromagnetic emission from compact binaries. This involved several advances in the \ac{GW} data analysis \citep{CBCLowLatency}, including the first real\nobreakdashes-time matched\nobreakdashes-filter detection pipeline (\acs{MBTA}; \aclu{MBTA}), a semi\nobreakdashes-coherent, ad hoc rapid triangulation code (Timing++), and the first version of a rigorous Bayesian \ac{MCMC} parameter estimation code (LALINFERENCE; \citealt{S6PE})---all in service of the first search for X\nobreakdashes-ray \citep{SwiftFollowup} and optical \citep{OpticalImageAnalysis} counterparts of \ac{GW} triggers, by a consortium of facilities. Despite the technical achievements in the \ac{GW} data analysis, there was an undesirable tradeoff between the speed of the rapid localization and the full parameter estimation: the former could analyze a detection candidate in minutes, whereas the latter took half a day; the latter decreased the area on the sky by a factor of 1/20 over the former but took 1000 times as long to run \cite{SiderySkyLocalizationComparison}.

The success of \ac{EM} follow\nobreakdashes-up of \ac{LIGO} events will depend critically on disseminating high quality sky localizations within a timescale of minutes to hours. To that end, we have devised a rapid and accurate Bayesian sky localization method that takes mere seconds to achieve approximately the same accuracy as the full \ac{MCMC} analysis. Our key insights are the following:
%
\begin{enumerate}
    \item Nearly all of the information in the \ac{GW} time series that is informative for sky localization is encapsulated within the matched\nobreakdashes-filter estimates of the times, amplitudes, and phases on arrival at the detectors. To infer the position and distance of a \ac{GW} event, we only have to consider three numbers per detector rather than a densely sampled strain time series per detector.
    \item The matched\nobreakdashes-filter pipeline can be treated as a measurement system in and of itself. Just like the strain time series from the detectors, the resultant times, amplitudes, and phases have a predictable and quantifiable measurement uncertainty that can be translated into a likelihood function suitable for Bayesian inference.
    \item Errors in sky localization are semi\nobreakdashes-independent from errors in masses. If we care only about position reconstruction and not about jointly estimating masses as well, then we can reduce the dimensionality of the parameter estimation problem significantly. Moreover, this frees us of the need to directly compute the expensive post-Newtonian model waveforms, making the likelihood itself much faster to evaluate.
    \item Thanks to a simple likelihood function and a well\nobreakdashes-characterized parameter space, we may dispense with costly and parallelization\nobreakdashes-resistant \ac{MCMC} integration, and instead perform the Bayesian marginalization with classic, deterministic, very low order Gaussian quadrature.
\end{enumerate}
%
We call the resulting algorithm \ac{BAYESTAR}\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \ac{GW} detectors may some day rise against us humans.}\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}. It is as fast as Timing++, but nearly as accurate as the rigorous full parameter estimation. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. Beginning with the first Advanced \acs{LIGO} observing run, \ac{BAYESTAR} will provide localizations within minutes of the detection of any \ac{BNS} merger candidate, playing a a key role in enabling rapid follow-up observations.

This paper is organized as follows. In Section~\ref{sec:detection}, we describe the \ac{GW} signal model and sketch the standard detection algorithm, the matched filter bank. In Section~\ref{sec:parameter-estimation}, we describe Bayesian inference formalism and the prevailing method for inferring the parameters of detected candidates, \ac{MCMC} sampling. In Section~\ref{sec:bayestar-likelihood}, we propose the \ac{BAYESTAR} likelihood as a model for the uncertainty in the matched\nobreakdashes-filter parameter estimates, and discuss its relationship to and consistency with the likelihood for the full \ac{GW} data. In Section~\ref{sec:prior}, we describe the input to \ac{BAYESTAR} supplied by the detection pipeline, and the prior distribution on parameters. In Section~\ref{sec:marginal-posterior}, we explain the integration scheme by which the posterior probability distribution is calculated for a given sky location. In Section~\ref{sec:adaptive-sampling}, we show a scheme whereby the sky posterior is sampled on an adaptive HEALPix grid. In Section~\ref{sec:run-time}, we report the running time of the algorithm on the hardware available on the \ac{LIGO} Data Grid. Finally, in Section~\ref{sec:case-study}, we quantify the sky localization performance on a comprehensive set of simulated \ac{GW} events.

% Establish and describe the first rapid and tractable solution of the GW position reconstruction problem.
% Use and abuse of Fisher matrix: use it as a clue.
% Generality: replace GW antenna pattern and GW polarization model with any other kind of receiver.
% Demonstration of a practical Bayesian analysis that does not use an MCMC method.

\section{Signal model and detection}
\label{sec:detection}

In the \ac{TD}, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}
    y_i(t) = x_i (t; \bm\theta) + n_i (t).
\end{equation}
%
In the \ac{FD},
%
\begin{equation}\label{eq:signal-model}
    Y_i (\omega) = \int_{-\infty}^\infty y(t) e^{-i \omega t} dt = X_i (\omega; \bm\theta) + N_i (\omega),
\end{equation}
%
where $X_i (\omega; \bm\theta)$ is the \ac{GW} signal given a parameter vector $\bm\theta$ that describes the \ac{GW} source, and $N_i (\omega)$ is that detector's Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[\left|N_i(\omega)\right|^2\right] + E\left[\left|N_i(-\omega)\right|^2\right] = 2 E\left[\left|N_i(\omega)\right|^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf Y (\omega) \equiv \{Y_i (\omega)\}_i$.

Under the assumptions that the detector noise is Gaussian and that the noise from different detectors are uncorrelated, the likelihood of the observation, $\mathbf y$, conditioned upon the parameters $\bm\theta$, is a product of Gaussian distributions:
%
\begin{multline}\label{eq:gaussian-likelihood}
    \mathcal{L}(\mathbf Y; \bm\theta) = \prod_i p(Y_i | \bm\theta)
        \\
        \propto \exp \left[
        - \frac{1}{2} \sum_i \int_0^\infty \frac{\left|Y_i (\omega)
            - X_i(\omega; \bm\theta) \right|^2}{S_i(\omega)} \, d\omega
    \right].
\end{multline}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation, and intrinsic parameters describing the physical properties of the binary components\footnote{This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it does play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\citep{PhysRevD.87.043004}, \ac{BNS} systems formed by binary stellar evolution should almost always circularize due to tidal interaction~\citep{0004-637X-572-1-407} and later \ac{GW} emission~\citep{PhysRev.136.B1224} long before the inspiral enters LIGO's frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\citep{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely: $\mathbf{S}_1 = \mathbf{S}_2 = 0$.}:
%
\begin{equation}\label{eq:params}
    \bm\theta = \left[
    \begin{array}{cl@{\quad}p{4cm}lp{3.75cm}}
        \alpha & \rdelim]{11}{0mm} & right ascension & \rdelim\}{7}{1mm} & \multirow{7}{2cm}{extrinsic parameters, $\bm\theta_\mathrm{ex}$} \\
        \delta && declination & \\
        r && distance & \\
        t_\oplus && arrival time at geocenter & \\
        \iota && inclination angle & \\
        \psi && polarization angle & \\
        \phi_c && coalescence phase & \\
        \cline{1-1}\cline{3-3}
        m_1 && first component's mass & \rdelim\}{4}{1mm} & \multirow{4}{2cm}{intrinsic parameters, $\bm\theta_\mathrm{in}$.}\\
        m_2 && second component's mass & \\
        \mathbf S_1 && first component's spin & \\
        \mathbf S_2 && second component's spin & \\
    \end{array}\right.
\end{equation}

Assuming a nonprecessing circular orbit, we can write the \ac{GW} signal received by any detector as a linear combination of two basis waveforms, $H_0$ and $H_{\pi/2}$ \citep{PhysRevD.83.084002}. $H_0$ and $H_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. If $H_0$ and $H_{\pi/2}$ are Fourier transforms of real functions, then $H_0(\omega) = H_0^*(-\omega)$ and $H_{\pi/2}(\omega) = H_{\pi/2}^*(-\omega)$, and we can write (assuming an arbitrary phase convention)
%
\begin{equation}
    H_{\pi/2}(\omega) = H_0(\omega) \cdot
    \begin{cases}
        -i & \text{if } \omega \geq 0 \\
        i & \text{if } \omega < 0
    \end{cases}.
\end{equation}
%
For brevity, we define $H \equiv H_{0}$ and write all subsequent equations in terms of the $H$ basis vector alone. Then, we can write the signal model in a way that isolates all dependence on the extrinsic parameters, $\bm\theta_\mathrm{ex}$, into a few coefficients, and all dependence on the intrinsic parameters, $\bm\theta_\mathrm{in}$, into the basis waveform, by taking the Fourier transform of Equation~(2.8) of \cite{PhysRevD.83.084002}:
%
\begin{multline}\label{eq:full-signal-model}
    X_i(\omega; \bm\theta) = e^{-i \omega (t_\oplus - \mathbf{d}_i \cdot \mathbf{n})}
    \frac{r_{1,i}}{r}
    e^{2 i \phi_c}
    \\
    \left[
    \frac{1}{2} \left(1 + \cos^2 \iota\right) \Re \left\{\zeta\right\} - i
    \left(\cos\iota\right) \Im \left\{\zeta\right\}
    \right]
    H(\omega; \bm\theta_\mathrm{in})
\end{multline}
%
for $\omega \geq 0$, where
%
\begin{equation}
    \zeta = e^{-2 i \psi} \left(
    F_{+,i}(\alpha, \delta, t_\oplus) +
    i F_{\times,i}(\alpha, \delta, t_\oplus)
    \right).
\end{equation}
%
The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, defined such that $0 \leq {F_{+,i}}^2 + {F_{\times,i}}^2 \leq 1$. They depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event and are presented in~\cite{ExcessPower}. In a coordinate system with the $x$ and $y$ axes aligned with the arms of a detector, the antenna pattern is given in spherical polar coordinates as
%
\begin{align}
    F_+ &= -\frac{1}{2}(1 + \cos^2 \theta) \cos{2\phi}, \\
    F_\times &= -\cos\theta \sin{2\phi}.
\end{align}
%
The unit vector $\mathbf{d}_i$ represents the position of detector $i$ in units of light travel time.\footnote{When considering transient \ac{GW} sources such as those that we are concerned with in this thesis, the origin of the coordinate system is usually taken to be the geocenter. For long\nobreakdashes-duration signals such as those from statically deformed neutron stars, the solar system barycenter is a more natural choice.} The vector $\mathbf{n}$ is the direction of the source. The negative sign in the dot product $-\mathbf{d}_i \cdot \mathbf{n}$ is present because the direction of travel of the \ac{GW} signal is opposite to that of its sky location. The quantity $r_{1,i}$ is a fiducial distance at which detector $i$ would register SNR=1 for an optimally oriented binary (face\nobreakdashes-on, and in a direction perpendicular to the interferometer's arms):
%
\begin{equation}\label{eq:horizon}
r_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|H(\omega; \boldsymbol \theta_\mathrm{in})\right|^2}{S_i(\omega)} \,d\omega.
\end{equation}

More succinctly, we can write the signal received by detector $i$ in terms of observable extrinsic parameters $\bm\theta_i = (\rho_i, \gamma_i, \tau_i)$, the amplitude $\rho_i$, phase $\gamma_i$, and time delay $\tau_i$ on arrival at detector $i$:
%
\begin{multline}\label{eq:signal-model}
    X_i (\omega; \bm\theta_i, \bm\theta_\mathrm{in})
    \\
    = X_i (\omega; \rho_i, \gamma_i, \tau_i, \bm\theta_\mathrm{in}) = \frac{\rho_i}{\sigma_i} e^{i (\gamma_i - \omega \tau_i)} H(\omega; \bm\theta_\mathrm{in}).
\end{multline}

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from the likelihood in Equation~(\ref{eq:gaussian-likelihood}) and the signal model in Equation~(\ref{eq:signal-model}). Concretely, this results in a bank of matched filters, or cross-correlations between the incoming data stream and a collection of template waveforms,
%
\begin{equation}
z_i(\tau_i;\bm\theta_\mathrm{in}) = \frac{1}{\sigma_i (\bm\theta_\mathrm{in})} \int_0^\infty \frac{H^*(\omega; \bm\theta_\mathrm{in}) Y_i(\omega) e^{i \omega \tau_i}}{S_i(\omega)} \,d\omega.
\end{equation}
%
The \ac{ML} point estimates of the signal parameters, $\mathrm{MLE}(\mathbf{y}) = \{\{ \hat{\bm\theta}_i \}_i, \hat{\bm\theta}_\mathrm{in}\} = \{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\bm\theta}_\mathrm{in}\}$, are given by
%
\begin{eqnarray}
    \label{eq:optimal-tau}
    \hat{\bm\theta}_\mathrm{in}, \{\hat\tau_i\}_i
        &=& \argmax_{\bm\theta_\mathrm{in}, \{\hat\tau_i\}_i}
        \sum_i \left| z_i\left(\tau_i;
        \bm\theta_\mathrm{in}\right) \right|^2, \\
    \label{eq:optimal-rho}
    \hat\rho_i &=& \left| z_i\left(\hat\tau_i;
        \hat{\bm\theta}_\mathrm{in}\right) \right|, \\
    \label{eq:optimal-gamma}
    \hat\gamma_i &=& \arg z_i\left(\hat\tau_i;
        \hat{\bm\theta}_\mathrm{in}\right).
\end{eqnarray}
%
A detection candidate consists of $\{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\bm\theta}_\mathrm{in}\}$. There are various ways to characterize the significance of a detection candidate. In Gaussian noise, the maximum likelihood for the network is obtained by maximizing the network \ac{SNR}, $\rho_\mathrm{net}$,
%
\begin{equation}
    \hat\rho_\mathrm{net} = \max_{\bm\theta} \sum_i {|z_i({\bm\theta})|}^2 = \sqrt{\sum_i {\hat\rho_i}^2};
\end{equation}
%
this, therefore, is the simplest useful candidate ranking statistic.

\section{Uncertainty and the Fisher matrix}
\label{sec:fisher}
